Linear stability of magnetized massive protoplanetary disks

### 2014-05-28 paper/paper.tex
first revision done

\begin{abstract}
  Magneto-rotational instability (MRI) and gravitational instability
  (GI) are the two principle routes to turbulent angular momentum
  transport in accretion disks. Protoplanetary disks may develop 
  both. This paper aims to reinvigorate interest in the study of
  magnetized  massive protoplanetary disks, starting from the basic
  issue of stability. The local linear stability of a 
  self-gravitating, uniformly magnetized, differentially rotating, 
  three-dimensional disk subject to axisymmetric perturbations is
  calculated numerically. The formulation includes 
  resistivity. It is found that the reduction in the disk thickness
  by self-gravity can decrease MRI growth rates; the MRI
  becomes global in the vertical direction, and MRI modes 
  with small radial length scales are stabilized. {\bf The
    upper limit to the vertical field strength to permit the MRI
    in a strongly self-gravitating disk is estimated.} In massive
  disks with layered resistivity, the MRI is not well-localized
  to regions where the Elsasser number exceeds unity. For MRI modes
  with radial length scales on the order of the disk thickness, self-gravity 
  can enhance density perturbations, an effect that becomes 
  significant in the presence of a strong toroidal field, {\bf and
    which depends on the symmetry of the underlying MRI mode.} In
  gravitationally unstable disks where GI and MRI growth rates are
  comparable, the character of unstable modes can transition smoothly
  between MRI and GI. Implications for non-linear simulations are
  discussed briefly.

>>>>

\begin{abstract}
  Magneto-rotational instability (MRI) and gravitational instability
  (GI) are the two principle routes to turbulent angular momentum
  transport in accretion disks. Protoplanetary disks may develop 
  both. This paper aims to reinvigorate interest in the study of
  magnetized  massive protoplanetary disks, starting from the basic
  issue of stability. The local linear stability of a 
  self-gravitating, uniformly magnetized, differentially rotating, 
  three-dimensional {\bf stratified} disk subject to axisymmetric perturbations is
  calculated numerically. The formulation includes 
  resistivity. It is found that the reduction in the disk thickness
  by self-gravity can decrease MRI growth rates; the MRI
  becomes global in the vertical direction, and MRI modes 
  with small radial length scales are stabilized. {\bf The
    upper limit to the vertical field strength to permit the MRI
    in a strongly self-gravitating disk is estimated.} In massive
  disks with layered resistivity, the MRI is not well-localized
  to regions where the Elsasser number exceeds unity. For MRI modes
  with radial length scales on the order of the disk thickness, self-gravity 
  can enhance density perturbations, an effect that becomes 
  significant in the presence of a strong toroidal field, {\bf and
    which depends on the symmetry of the underlying MRI mode.} In
  gravitationally unstable disks where GI and MRI growth rates are
  comparable, the character of unstable modes can transition smoothly
  between MRI and GI. Implications for non-linear simulations are
  discussed briefly.

### 2014-05-28 paper/results_azi.tex
first revision done

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2cm 0cm
    0cm]{figures/compare_growth3_tilted_even.ps}  
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
    0.52cm]{figures/compare_growth3_tilted_odd.ps} 
  \caption{MRI growth rates in isothermal disks with $Q=~0.2$ ($Q_\mathrm{2D}=0.72$) and 
    $\beta=10$ for a range of azimuthal field strengths $B_y/B_z$. The
    dots are solutions computed from the full problem, with the
    colorbar measuring the gravitational potential perturbation, while
    the solid curves are computed from the Cowling approximation. 
    For $B_y=0$, modes in top and bottom panels have
    $W^\prime(0)=0$ and $W(0)=0$, respectively.
    \label{compare_growth3_tilted}}
\end{figure}

@@

The modes in the bottom panel of Fig. \ref{compare_growth3_tilted} 
display opposite behavior. For $B_y=0$ we find $W$ is odd, and
self-gravity has no effect. When $B_y>0$, $W$ deviates from an
odd function and the midplane density perturbation $|W(0)|$
increases. SG is stabilizing for these modes at all wavelengths, and
is most effective at $k_xH =  O(1)$.

Fig. \ref{result_tilted} show eigenfunctions for $\epsilon=3$ and
$k_xH=1.1$ with and without the Cowling approximation.
SG significantly enhances the midplane density perturbation, making the
gravitational potential energy comparable to the magnetic energy, 
which becomes more confined near the midplane.

@@

Notice also in Fig. \ref{compare_growth3_tilted_resis} the disappearance of
magnetic modes between $0.8\lesssim k_xH\lesssim 1.5$ as $B_y$ is
increased. For $B_y=2B_z$, MRI and GI are again independent, but
because they operate at distinct radial scales. This implies that
perturbations unstable to GI cannot develop MRI.

>>>>

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2cm 0cm
    0cm]{figures/compare_growth3_tilted_even.ps}  
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
    0.52cm]{figures/compare_growth3_tilted_odd.ps} 
  \caption{MRI growth rates in isothermal disks with $Q=~0.2$ ($Q_\mathrm{2D}=0.72$) and 
    $\beta=10$ for a range of azimuthal field strengths $B_y/B_z$. The
    dots are solutions computed from the full problem, with the
    colorbar measuring the gravitational potential perturbation {\bf
      via $\tau$}, while
    the solid curves are computed from the Cowling approximation. 
    For $B_y=0$, modes in top and bottom panels have
    $W^\prime(0)=0$ and $W(0)=0$, respectively.
    \label{compare_growth3_tilted}}
\end{figure}

@@

The modes in the bottom panel of Fig. \ref{compare_growth3_tilted} 
display opposite behavior. For $B_y=0$ we find {\bf$W(z)$} is odd, and
self-gravity has no effect. When $B_y>0$, $W$ deviates from an
odd function and the midplane density perturbation $|W(0)|$
increases. SG is stabilizing for these modes at all wavelengths, and
is most effective at $k_xH =  O(1)$. 
Fig. \ref{result_tilted} show eigenfunctions for $\epsilon=3$ and
$k_xH=1.1$ with and without the Cowling approximation.
SG significantly enhances the midplane density perturbation, making the
gravitational potential energy comparable to the magnetic energy, 
which becomes more confined near the midplane.

@@

Notice also in Fig. \ref{compare_growth3_tilted_resis} the disappearance of
magnetic modes between $0.8\lesssim k_xH\lesssim 1.5$ as $B_y$ is
increased. For $B_y=2B_z$, MRI and GI are again independent 
because they operate at distinct radial scales. This implies that
perturbations unstable to GI cannot develop MRI.

### 2014-05-28 paper/results_gi.tex
revision to 1st ref report
first revision done

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2.1cm 0cm
    0cm]{figures/compare_growth3_kx1_Q0d18.ps} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2.1cm 0cm
    0.5cm]{figures/compare_growth3_kx1_Q0d14.ps} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
    0.5cm]{figures/compare_growth3_kx1_Q0d12.ps} 
  \caption{Growth rates for modes with $k_xH=1$ in isothermal ideal
    disks with $Q=0.18$ (top), $Q=0.14$ (middle) and $Q=0.14$ (bottom). The
    colorbar measures the importance of  
    self-gravity by $\tau$.   
    \label{compare_growth3}}
\end{figure}

The GI and MRI branches only interact when their growth rates
are similar. This is seen in Fig. \ref{compare_growth3} for $Q=~0.18$
where the GI branch approaches a MRI branch at $\beta\simeq 25,\,
\gamma\simeq 0.2\Omega$. In fact, following the red curve to smaller
$\beta$ indicates GI transitions to MRI. The
`gaps' in the GI and MRI branches for $Q=0.18$ and $Q=0.12$ may be due
to the phenomenon of avoided crossing, as seen in stars
\citep[e.g.][]{aizenman77} and accretion tori
\citep[e.g.][]{christo93}, where physically distinct modes approach
one another in frequency and exchange character. However, we cannot
exclude the possibility  that some modes may have been missed in a 
numerical search of eigenfrequencies.  
%maybe ogilvie?

@@

The dispersion relation $\gamma(k_x)$ deviates from the
pure GI/MRI curves with increasing $k_x$, implying stronger interaction between
magnetic and density perturbations. Comparing pure GI (dashed line)
and the gravitationally-dominated portions of $\gamma(k_x)$ shows that 
inclusion of magnetic field stabilizes high-$k_x$ pure GI. (Note also
the slight decrease in the most unstable $k_x$.) 
This stabilization is due to magnetic pressure \citep{lizano10}, which
is consistent with stabilizing the small wavelength GI only.

Comparing pure MRI (solid line) and the magnetically-dominated
portions of $\gamma(k_x)$ show that self-gravity increases MRI growth
rates at large $k_x$. This effect is small but noticeable, which can
be used as a code test for non-linear simulations. Note that this
destabilization is through the linear response, rather 
than through the background stratification (which is stabilizing).

>>>>

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2.1cm 0cm
    0cm]{figures/compare_growth3_kx1_Q0d18.ps} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2.1cm 0cm
    0.5cm]{figures/compare_growth3_kx1_Q0d14.ps} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
    0.5cm]{figures/compare_growth3_kx1_Q0d12.ps} 
  \caption{Growth rates for modes with $k_xH=1$ in isothermal ideal
    disks with $Q=0.18$ (top), $Q=0.14$ (middle) and $Q=0.12$ (bottom). The
    colorbar measures the importance of  
    self-gravity by $\tau$.   
    \label{compare_growth3}}
\end{figure}

The GI and MRI branches only interact when their growth rates
are similar. This is seen in Fig. \ref{compare_growth3} for $Q=~0.18$
where the GI branch approaches a MRI branch at $\beta\simeq 25,\,
\gamma\simeq 0.2\Omega$. In fact, following the red curve to smaller
$\beta$ indicates GI transitions to MRI. The
`gaps' in the GI and MRI branches for $Q=0.18$ and $Q=0.12$ may be due
to the phenomenon of avoided crossing, as seen in stars
\citep[e.g.][]{aizenman77} and accretion tori{\bf /disks
\citep[e.g.][]{christo93,ogilvie98}}, where physically distinct modes approach
one another in frequency and exchange character. However, we cannot
exclude the possibility  that some modes may have been missed in a 
numerical search of eigenfrequencies.  
%maybe ogilvie?

@@

The dispersion relation $\gamma(k_x)$ deviates from the
pure GI/MRI curves with increasing $k_x$, implying stronger interaction between
magnetic and density perturbations. Comparing pure GI (dashed line)
and the gravitationally-dominated portions of $\gamma(k_x)$ shows that 
inclusion of magnetic field stabilizes high-$k_x$ pure GI. (Note also
the slight decrease in the most unstable $k_x$.) 
This stabilization is due to magnetic pressure \citep{lizano10}, {\bf 
consistent with pressure stabilizing small-wavelength GI only.}

Comparing pure MRI (solid line) and the magnetically-dominated
portions of $\gamma(k_x)$ show that self-gravity increases MRI growth
rates at large $k_x$. This effect is small but noticeable, which can
be used as a code test for non-linear simulations. Note that this
destabilization is through {\bf self-gravity in} the linear response, rather 
than through the background stratification (which is stabilizing).

### 2014-05-28 paper/summary.tex
first revision done

For a vertical field, the requirement for the MRI {\bf to operate} is that   
its vertical wavelength $\lambda \lesssim 2H$. 
{\bf 
%While $\lambda$ is independent of
%the strength of self-gravity $Q$ {\bf for fixed $\beta$}, 
The disk thickness $H=H(Q)$ decreases with increasing SG. This can
reduce MRI growth rates when 
$\beta$, and hence $\lambda$, is fixed.  
}
Thus, a sufficiently massive disk can potentially suppress the MRI. 
The MRI is also restricted to larger radial 
scales as $Q$ is lowered. This means that the MRI becomes more global
in self-gravitating disks.

{\bf
  The condition $\lambda < 2H$ 
  %requires $f(Q)/\sqrt{\beta}$ to
  %  be sufficiently small (see
  %  Eq. \ref{lambda_ideal}---\ref{lambda_resis}), so larger values of
  %  $\beta$ are required to  
  %  permit the MRI with increasing self-gravity, since $f(Q)$ increases
  %  with decreasing $Q$. 
  may be written more precisely as 
  \begin{align}\label{cond1}
    \frac{n}{\mathrm{min}(\Lambda_0, 1)}\frac{f(Q)}{\sqrt{\beta}} \lesssim 1,
  \end{align}
  where $n\sim 3$ is a numerical factor and
  $\mathrm{min}(\Lambda_0,1)$ accounts for the ideal and resistive
  limits (see Eq. \ref{lambda_ideal}---\ref{lambda_resis}). Since $f$
  increases with decreasing $Q$, Eq. \ref{cond1} implies
  the MRI requires larger $\beta$ values with  
  increasing self-gravity. 
  For definiteness, consider the ideal polytropic disk. Then Eq. \ref{cond1} is
  \begin{align}\label{cond2}
    \beta^{-1/2} \lesssim
    \frac{\sqrt{15}}{4\pi}\sqrt{Q}\arccos{\left(\frac{Q}{1+Q}\right)}. 
  \end{align}
  For a non-self-gravitating disk, $Q\to\infty$ and Eq. \ref{cond2} is
  $\beta \gtrsim 16\pi^2/30\simeq 5$. For $Q\ll 1$, the condition is
  $\beta \gtrsim 64/15Q$, giving $\beta\gtrsim 20$ for
  $Q=0.2$. We confirmed this numerically, finding that the growth rate
  $\gamma  \lesssim 0.1$ when $\beta \lesssim 3.3$ for $Q=20$ and
  $\beta \lesssim 17$ for $Q=0.2$.  
%  Inserting $Q\sim 0.2$ for an ideal disk marginally stable to
%  self-gravity, we find $v_{A0}\lesssim 0.2\csmid$ for the MRI.

  We can also place an upper bound on the field strength
  $B_z$. Writing $v_{A0} = B_z\sqrt{4\pi G Q/\mu_0\Omega^2}$, we find 
  Eq. \ref{cond2} is independent of $Q$ for $Q\ll1$, and 
  \begin{align}\label{cond3}
    \frac{B_z}{\csmid\Omega}\sqrt{\frac{\pi G}{\mu_0}}\lesssim
    \frac{\sqrt{15}}{16}
  \end{align}
  is needed for the MRI to operate in an ideal polytropic disk with strong
  self-gravity. Although both the MRI wavelength and disk thickness
  vanish as $Q\to 0$, the MRI can still operate provided the field is
  sufficiently weak according to Eq. \ref{cond3}.

>>>>

For a vertical field, the requirement for the MRI {\bf to operate} is that   
its vertical wavelength $\lambda \lesssim 2H$. 
{\bf 
%While $\lambda$ is independent of
%the strength of self-gravity $Q$ {\bf for fixed $\beta$}, 
The disk thickness $H=H(Q)$ decreases with increasing SG. This 
reduces MRI growth rates when 
$\beta$, and hence $\lambda$, is fixed.  
}
Thus, a sufficiently massive disk can potentially suppress the MRI. 
The MRI is also restricted to larger radial 
scales as $Q$ is lowered. This means that the MRI becomes more global
in self-gravitating disks.

{\bf
  The condition $\lambda < 2H$ 
  %requires $f(Q)/\sqrt{\beta}$ to
  %  be sufficiently small (see
  %  Eq. \ref{lambda_ideal}---\ref{lambda_resis}), so larger values of
  %  $\beta$ are required to  
  %  permit the MRI with increasing self-gravity, since $f(Q)$ increases
  %  with decreasing $Q$. 
  may be written more precisely as 
  \begin{align}\label{cond1}
    \frac{n}{\mathrm{min}(\Lambda_0, 1)}\frac{f(Q)}{\sqrt{\beta}} \lesssim 1,
  \end{align}
  where $n\sim 3$ is a numerical factor and
  $\mathrm{min}(\Lambda_0,1)$ accounts for the ideal and resistive
  limits (see Eq. \ref{lambda_ideal}---\ref{lambda_resis}). Since $f$
  increases with decreasing $Q$, Eq. \ref{cond1} implies
  the MRI requires larger values of $\beta$ with  
  increasing self-gravity. 
  For definiteness, consider the ideal polytropic disk. Then Eq. \ref{cond1} is
  \begin{align}\label{cond2}
    \beta^{-1/2} \lesssim
    \frac{\sqrt{15}}{4\pi}\sqrt{Q}\arccos{\left(\frac{Q}{1+Q}\right)}. 
  \end{align}
  For a non-self-gravitating disk, $Q\to\infty$ and Eq. \ref{cond2} is
  $\beta \gtrsim 16\pi^2/30\simeq 5$. For $Q\ll 1$, the condition is
  $\beta \gtrsim 64/15Q$, giving $\beta\gtrsim 20$ for
  $Q=0.2$. We confirm this numerically, finding the MRI growth rate
  $\gamma  \lesssim 0.1$ when $\beta \lesssim 3.3$ for $Q=20$ and 
  $\beta \lesssim 17$ for $Q=0.2$.  
%  Inserting $Q\sim 0.2$ for an ideal disk marginally stable to
%  self-gravity, we find $v_{A0}\lesssim 0.2\csmid$ for the MRI.

  We can also place an upper bound on the absolute field strength
  $B_z$. Writing $v_{A0} = B_z\sqrt{4\pi G Q/\mu_0\Omega^2}$, we find 
  Eq. \ref{cond2} is independent of $Q$ for $Q\ll1$, and 
  \begin{align}\label{cond3}
    \frac{B_z}{\csmid\Omega}\sqrt{\frac{\pi G}{\mu_0}}\lesssim
    \frac{\sqrt{15}}{16}
  \end{align}
  is needed for the MRI to operate in the ideal polytropic disk with strong
  self-gravity. Although both the MRI wavelength and disk thickness
  vanish as $Q\to 0$, the MRI can still operate provided the field is
  sufficiently weak according to Eq. \ref{cond3}.
